Concurrent Programming in Harmony

### 2021-12-30 book/paper.tex
Latest

@@

\autoref{fig:tas} goes on to implement mutual exclusion for
a set of <{N}> threads.
The approach is called \emph{spinlock},
\index{spinlock}%
because each thread is ``spinning'' (executing a tight loop) until
it can acquire the lock.
The program uses $<{N}>+1$ boolean variables.
Variable <{shared}> is initialized to <{False}> while
<{private}[i]}> for each thread <{i}> is initialized to <{True}>.

@@

To see that invariant $\mathcal{I}_1$ is maintained, note that
<{!p = True}> upon entry of <{test_and_set}>
(because of the condition on the <{while}> loop that the
<{test_and_set}> method is invoked in).
There are two cases:
\begin{enumerate}
\item <{!s}> is <{False}> upon entry to <{test_and_set}>.
Then upon exit <{!p = False}> and <{!s = True}>,
maintaining the invariant.
\item <{!s}> is <{True}> upon entry to <{test_and_set}>.
Then upon exit nothing has changed, maintaining the invariant.
\end{enumerate}
Invariant $\mathcal{I}_1$ is also easy to verify for exiting the critical section
because we can assume, by the induction hypothesis, that $<{private}>[i]$ is
<{True}> just before exiting the critical section.
Invariant $\mathcal{I}_2(i)$ is obvious as (i) thread <{i}> only proceeds to the critical
section if <{private}>[<{i}>] is <{False}>, and (ii) no other thread modifies
<{private}>[<{i}>].

@@

\autoref{fig:qtestpar} also shows the deterministic finite automaton
that describes the possible outputs when the test
program is run against the specification in the case
$<{NOPS}> = 2$---for $<{NOPS}> = 4$ it would be much
too large to print here.  Since there are no cycles in the DFA, you
can follow some paths through this DFA and see that they are valid
interleavings of the threads.
You can obtain this output yourself by running

@@

You can
try this for various $<{NOPS}>$, although it gets increasingly
harder to check by hand that the generated DFAs are the same as
<{NOPS}> increases.
Now run the test program against \autoref{fig:queueseq}, which is
clearly an incorrect implementation of the concurrent queue specification
because it contains no synchronization code.
However, Harmony does not immediately detect
any problems.  In fact, for $\mathtt{NOPS} = 2$ it even generates
the same set of behaviors.
This is because the test program only outputs the behaviors---it does
not check if they are correct.

@@

\begin{figure}
\begin{center}
\includegraphics[width=\textwidth]{figures/abdtest.png}
\end{center}
\begin{code}
<{:input:../code/abdtest.hny}>
\end{code}
\caption{\harmonylink{code/abdtest.hny} Testing the code in \autoref{fig:abd}}
\label{fig:abdtest}
\end{figure}

@@

A \emph{register} is an object that you can read or write.  In a distributed
system, it can be useful to have registers that are shared.
A simple shared register implementation
would have its value maintained by a server, and clients can read or write the
shared register by exchanging messages with the server.
We call two operations such that one does not finish before the other starts
\emph{concurrent}.
Since messages are delivered one at a time to the server,
concurrent operations on the shared register appear atomic.
In particular, we have the following three desirable properties:
\begin{enumerate}
\item All write operations are ordered;
\item A read operation returns either the last value written or the value of
a concurrent write operation.
\item If read operation $r_1$ finishes before read operation $r_2$ starts,
then $r_2$ cannot return a value that is older than the value returned
by $r_1$.
\end{enumerate}

\index{single point of failure}%
Unfortunately, however, the server is a \emph{single point of failure}:
if it fails, all its clients suffer.
We would therefore like to find a solution that can survive the crash
of a server.
While we could use Chain Replication to replicate the register, in
this chapter we will use a solution that does not assume that
crashes can be accurately detected.

@@

\index{logical timestamp}%
We present a protocol preserving these properties that is based on the
work by Hagit Attiya, Amotz Bar-Noy, and Danny Dolev~\cite{ABD95}.
In order to tolerate <{F}> failures,
it uses $<{N}> = 2<{F}> + 1$ servers.
In other words, the register survives as long as a strict majority of
its copies survive.
All write operation will be ordered by a unique \emph{logical timestamp}
(see also \autoref{ch:testing}).
Each server maintains not only the value of the object, but also the
timestamp of its corresponding write operation.

Each read and write operation consists of two \emph{phases}.  In a phase,
a client broadcasts a request to all servers and waits for responses
from a majority ($<{N}> - <{F}>$ or equivalently
$<{F}> + 1$ servers).  Note that because we are assuming that no
more than <{F}> servers can fail, doing so is safe, in that a client
cannot indefinitely block as long as that assumption holds.

In the first phase, a client asks each server for its current
timestamp and value.  After receiving $<{N}> - <{F}>$ responses,
the client determines the response with the highest timestamp.
In case of a write operation, the client then computes a new unique
timestamp that is strictly higher than the highest it has seen.
To make this work, timestamps are actually lexicographically ordered
tuples consisting of an integer and the unique identifier of the
client that writes the value.  So, if $(t, c)$ is the highest timestamp
observed by client $c'$, and $c'$  needs to create a new timestamp,
it can select $(t + 1, c')$.
After all $(t + 1, c') > (t, u)$ and no other client
will create the same timestamp.

Suppose client $c'$ is trying to write a value <{v}>.
In phase 2, client $c'$ broadcasts a request containing timestamp $(t+1, c')$
and <{v}>.  Each server that receives the request compares $(t+1, c')$ against
its current timestamp.  If $(t+1, c')$ is larger than its current timestamp,
it adopts the new timestamp and its corresponding value <{v}>.  In either case,
the server responds to the client.
Upon $c'$ receiving a response from $<{N}> - <{F}>$ servers,
the write operation completes.
In case of a read operation, client $c'$ simply \emph{writes back} the
highest timestamp it saw in the first phase along with its corresponding value.

@@

Read and write operations are both invoked with a unique identifier
<{uid}>.  Both start by broadcasting a <{.read}>
request to all servers and then waiting for a response from
<{N - F}> servers.
The <{receive()}> function uses the <{bag.combinations}> method
to find all combinations of subsets of responses of size
$<{N}> - <{F}>$.
The second phase of each operation is similar.

\autoref{fig:abdtest} tests the third property of the properties listed
above.  The <{writer()}> method writes a new value <{.token}>.
The \texttt{reader()} method reads twice and makes sure that if the
first read operation returns <{.token}>, then so does the second.
This illustrates the importance of the second phase of the \texttt{read}
operation.  You can comment out Lines~32 and~33 in \autoref{fig:abd} and
see what goes wrong.

One may wonder how failures can occur in this model.  They are not
explicitly modeled, but Harmony tries every possible execution.
This includes executions in which the clients terminate before
$<{F}>$ of the servers start executing.  To the clients, this
is indistinguishable from executions in which those
servers have failed.

@@

This is just the specification---in practice we do not have a shared
variable in which we can store the decision a priori.
We will present a simple consensus algorithm that can tolerate
fewer than $1/3^{rd}$ of processors failing by crashing.
More precisely,
constant <{F}> contains the maximum number of failures, and
we will assume there are $<{N}> = 3<{F}> + 1$ processors.

@@

\autoref{fig:bosco} presents our algorithm.  Besides the <{network}>
variable,
it uses a shared list of proposals and a shared set of decisions.
In this particular algorithm, all messages are broadcast to all processors, so they do
not require a destination address.
The $<{N}>$ processors go through a sequence of \emph{rounds} in which they wait for
<{N - F}> messages, update their state based on the messages, and
broadcast messages containing their new state.
The reason that a processor waits for <{N - F}> rather than <{N}>
messages is because of failures: up to <{F}> processors may never send a message
and so it would be unwise to wait for all <{N}>.  You might be tempted to use
a timer and time out on waiting for a particular processor.  But how would you initialize
that timer?  While we will assume that the network is reliable, there is no guarantee
that messages arrive within a particular time.
We call a set of <{N - F}> processors a \emph{quorum}.
A quorum must suffice for the algorithm to make progress.

@@

The body of the <{atomically when exists}> statement contains the core of the algorithm.
Note that $<{N}> - <{F}> = 2<{F}> + 1$, so that the number of
messages is guaranteed to be odd.  Also, because there are only 0 and 1 values, there
must exist a majority of zeroes or ones.  Variable <{count[0]}> stores the number
of zeroes and <{count[1]}> stores the number of ones received in the round.
The rules of the algorithm are simple:
\begin{itemize}
\item update <{estimate}> to be the majority value;
\item if the quorum is unanimous, decide the value.
\end{itemize}
After that, proceed with the next round.

@@

Paxos~\cite{Paxos} is the most well-known family of consensus protocols for
environments in which few or no assumptions are made about timing.
In this chapter we present a basic version of a Paxos protocol, one that is
\emph{single-decree} (only tries to make a single decision).
It uses two kinds of processors: \emph{leaders} and \emph{acceptors}.
In order to tolerate <{F}> crash failures, you need at least $<{F}> + 1$ leaders
and $2<{F}> + 1$ acceptors, but leaders and acceptors can be colocated, so
in total only $2<{F}> + 1$ independently failing processors are needed.
Here we provide only a rudimentary introduction to Paxos; for more detailed
information refer to~\cite{Paxos}.

>>>>

% \renewcommand{\dblfloatpagefraction}{0.99}
% \renewcommand{\floatpagefraction}{.99}
% \renewcommand{\dblfloatpagefraction}{0.99}
% \renewcommand{\floatpagefraction}{.99}

@@

\autoref{fig:tas} goes on to implement mutual exclusion for
a set of <{N}> threads.
The approach is called \emph{spinlock},
\index{spinlock}%
because each thread is ``spinning'' (executing a tight loop) until
it can acquire the lock.
The program uses <{N + 1}> boolean variables.
Variable <{shared}> is initialized to <{False}> while
<{private}[i]}> for each thread <{i}> is initialized to <{True}>.

@@

To see that invariant $\mathcal{I}_1$ is maintained, note that
<{!p = True}> upon entry of <{test_and_set}>
(because of the condition on the <{while}> loop that the
<{test_and_set}> method is invoked in).
There are two cases:
\begin{enumerate}
\item <{!s}> is <{False}> upon entry to <{test_and_set}>.
Then upon exit <{!p = False}> and <{!s = True}>,
maintaining the invariant.
\item <{!s}> is <{True}> upon entry to <{test_and_set}>.
Then upon exit nothing has changed, maintaining the invariant.
\end{enumerate}
Invariant $\mathcal{I}_1$ is also easy to verify for exiting the critical section
because we can assume, by the induction hypothesis, that <{private[i]}> is
<{True}> just before exiting the critical section.
Invariant $\mathcal{I}_2(i)$ is obvious as (i) thread <{i}> only proceeds to the critical
section if <{private}>[<{i}>] is <{False}>, and (ii) no other thread modifies
<{private}>[<{i}>].

@@

\autoref{fig:qtestpar} also shows the deterministic finite automaton
that describes the possible outputs when the test
program is run against the specification in the case
<{NOPS = 2}>---for <{NOPS = 4}> it would be much
too large to print here.  Since there are no cycles in the DFA, you
can follow some paths through this DFA and see that they are valid
interleavings of the threads.
You can obtain this output yourself by running

@@

You can
try this for various <{NOPS}>, although it gets increasingly
harder to check by hand that the generated DFAs are the same as
<{NOPS}> increases.
Now run the test program against \autoref{fig:queueseq}, which is
clearly an incorrect implementation of the concurrent queue specification
because it contains no synchronization code.
However, Harmony does not immediately detect
any problems.  In fact, for $\mathtt{NOPS} = 2$ it even generates
the same set of behaviors.
This is because the test program only outputs the behaviors---it does
not check if they are correct.

@@

\begin{figure}
\begin{center}
\includegraphics[width=\textwidth]{figures/abdtest.png}
\end{center}
\begin{code}
<{:input:../code/abdtest.hny}>
\end{code}
\caption{\harmonylink{code/abdtest.hny} Behavioral test for atomic read/write registers and the output for the case that <{NREADERS = NWRITERS = 1}>}
\label{fig:abdtest}
\end{figure}

@@

A \emph{register} is an object that you can read or write.
In a distributed system, examples include a shared file system
(each file is a register) or a key/value store (each key corresponds
to a register).
A simple shared register implementation
would have its value maintained by a server, and clients can read or write the
shared register by exchanging messages with the server.
We call two operations such that one does not finish before the other starts
\emph{concurrent}.
Since messages are delivered one at a time to the server,
concurrent operations on the shared register appear atomic.
In particular, we have the following three desirable properties:
\begin{enumerate}
\item All write operations are ordered;
\item A read operation returns either the last value written or the value of
a concurrent write operation.
\item If read operation $r_1$ finishes before read operation $r_2$ starts,
then $r_2$ cannot return a value that is older than the value returned
by $r_1$.
\end{enumerate}

It is instructive to look at the test program and its output in
\autoref{fig:abdtest}.  This is for the case when there is only
a single reader thread (identified as ``1'') and a single writer
thread (identified as ``$-1$''), but already
there are many cases to consider.  Each thread prints information
just before and just after doing their single operation.  The output
shows all possible interleavings in the form of a DFA.  Note
that if the read operation starts after the write operation
has completed, then the read operaion must return the new value.
However, if the two operations interleave in some way, then the
read operation can return either the old or the new value.

\index{single point of failure}%
Unfortunately, a server is a \emph{single point of failure}:
if it fails, all its clients suffer.
We would therefore like to find a solution that can survive the crash
of a server.
While we could use Chain Replication to replicate the register, in
this chapter we will use a solution that does not assume that
crashes can be accurately detected.

@@

\index{logical timestamp}%
We present a protocol preserving these properties that is based on the
work by Hagit Attiya, Amotz Bar-Noy, and Danny Dolev~\cite{ABD95}.
In order to tolerate <{F}> failures,
it uses <{N}> = 2<{F}> + 1 servers.
In other words, the register survives as long as a strict majority of
its copies survive.
All write operation will be ordered by a unique \emph{logical timestamp}
(see also \autoref{ch:testing}).
Each server maintains not only the value of the object, but also the
timestamp of its corresponding write operation.

Each read and write operation consists of two \emph{phases}.  In a phase,
a client broadcasts a request to all servers and waits for responses
from a majority (<{N - F}> or equivalently
<{F + 1}> servers).  Note that because we are assuming that no
more than <{F}> servers can fail, doing so is safe, in that a client
cannot indefinitely block as long as that assumption holds.

In the first phase, a client asks each server for its current
timestamp and value.  After receiving <{N - F}> responses,
the client determines the response with the highest timestamp.
In case of a write operation, the client then computes a new unique
timestamp that is strictly higher than the highest it has seen.
To make this work, timestamps are actually lexicographically ordered
tuples consisting of an integer and the unique identifier of the
client that writes the value.  So, if $(t, c)$ is the highest timestamp
observed by client $c'$, and $c'$  needs to create a new timestamp,
it can select $(t + 1, c')$.
After all $(t + 1, c') > (t, u)$ and no other client
will create the same timestamp.

Suppose client $c'$ is trying to write a value <{v}>.
In phase 2, client $c'$ broadcasts a request containing timestamp $(t+1, c')$
and <{v}>.  Each server that receives the request compares $(t+1, c')$ against
its current timestamp.  If $(t+1, c')$ is larger than its current timestamp,
it adopts the new timestamp and its corresponding value <{v}>.  In either case,
the server responds to the client.
Upon $c'$ receiving a response from <{N - F}> servers,
the write operation completes.
In case of a read operation, client $c'$ simply \emph{writes back} the
highest timestamp it saw in the first phase along with its corresponding value.

@@

Read and write operations are both invoked with a unique identifier
<{uid}>.  Both start by broadcasting a <{.read}>
request to all servers and then waiting for a response from
<{N - F}> servers.
The <{receive()}> function uses the <{bag.combinations}> method
to find all combinations of subsets of responses of size <{N - F}>.
The second phase of each operation is similar.

\autoref{fig:abdtest} can be used to test this protocol, although
you will notice that the model checker cannot deal with cases involving
more than three client threads.
Three is just enough to check the third property listed above
(using one writer and two readers).
Doing so illustrates the importance of the second phase of the \texttt{read}
operation.  You can comment out Lines~34, 36, and~37 in \autoref{fig:abd}
and to elide the second phase and see what goes wrong.

One may wonder how failures can occur in this model.  They are not
explicitly modeled, but Harmony tries every possible execution.
This includes executions in which the clients terminate before
<{F}> of the servers start executing.  To the clients, this
is indistinguishable from executions in which those
servers have failed.

@@

This is just the specification---in practice we do not have a shared
variable in which we can store the decision a priori.
We will present a simple consensus algorithm that can tolerate
fewer than $1/3^{rd}$ of processors failing by crashing.
More precisely,
constant <{F}> contains the maximum number of failures, and
we will assume there are <{N}> = 3<{F}> + 1 processors.

@@

\autoref{fig:bosco} presents our algorithm.  Besides the <{network}>
variable,
it uses a shared list of proposals and a shared set of decisions.
In this particular algorithm, all messages are broadcast to all processors, so they do
not require a destination address.
The <{N}> processors go through a sequence of \emph{rounds} in which they wait for
<{N - F}> messages, update their state based on the messages, and
broadcast messages containing their new state.
The reason that a processor waits for <{N - F}> rather than <{N}>
messages is because of failures: up to <{F}> processors may never send a message
and so it would be unwise to wait for all <{N}>.  You might be tempted to use
a timer and time out on waiting for a particular processor.  But how would you initialize
that timer?  While we will assume that the network is reliable, there is no guarantee
that messages arrive within a particular time.
We call a set of <{N - F}> processors a \emph{quorum}.
A quorum must suffice for the algorithm to make progress.

@@

The body of the <{atomically when exists}> statement contains the core of the algorithm.
Note that <{N - F}> = 2<{F}> + 1, so that the number of
messages is guaranteed to be odd.  Also, because there are only 0 and 1 values, there
must exist a majority of zeroes or ones.  Variable <{count[0]}> stores the number
of zeroes and <{count[1]}> stores the number of ones received in the round.
The rules of the algorithm are simple:
\begin{itemize}
\item update <{estimate}> to be the majority value;
\item if the quorum is unanimous, decide the value.
\end{itemize}
After that, proceed with the next round.

@@

Paxos~\cite{Paxos} is the most well-known family of consensus protocols for
environments in which few or no assumptions are made about timing.
In this chapter we present a basic version of a Paxos protocol, one that is
\emph{single-decree} (only tries to make a single decision).
It uses two kinds of processors: \emph{leaders} and \emph{acceptors}.
In order to tolerate <{F}> crash failures, you need at least <{F}> + 1 leaders
and $2<{F}> + 1$ acceptors, but leaders and acceptors can be colocated, so
in total only $2<{F}> + 1$ independently failing processors are needed.
Here we provide only a rudimentary introduction to Paxos; for more detailed
information refer to~\cite{Paxos}.
